On the spatial aggregation of ecosystem condition indicators

### 2025-02-20 manuscript/manuscript.qmd
clearpage

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
  - aggregation error
  - aggregation displacement
  - upscaling
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
callout-appearance: simple
callout-icon: false
---

@@

@@

::: {.callout-note}
**Box 1 -- Related concepts**\

@@

It is common in the ECA community of practice to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e. horizontal) aggregation. \

@@

The modifiable area unit problem (MAUP) describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
Administrative units are often used as the ecosystem assets in SEEA EAs, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
The original spatial units that data are collected at may be a pragmatic choice in much the same way.
Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here. \

@@

```{=html}
<!-- The indicator arguably has much less room for miss-interpretation (relative to its intended use).
Although one common misunderstanding is that indicators named by an animal (e.g. wolves; Example 2) is taken by untrained users to represent a population status (typically the number of individuals).
This would, however, imply that the spatial aggregation function used for the indicator is summation, 
but the recommended, on arguably the most used function, is an area weighted mean.
With this aggregation function, indicators will always refer to the mean indicator value for all the areas considered, and if properly defined, refer to the degree of fulfillment for the ecosystem characteristic.
Sometimes changing the indicator name, or redefining the indicator, goes a long way towards easing the interpretation and avoiding miss-use (e.g. *mean reindeer densities*). -->
```

@@

```{r read}
#| include: false
dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
  mutate(X = X |> str_replace_all("%", "\\\\%"))
cn <- colnames(dat) |>
    str_replace_all("2", "") |>
    str_replace_all("X", "") |>
    str_replace_all("RL", "X\\\\textsubscript{100}")
```

```{r tbl-pti}
#| echo: false
#| tbl-cap: "Dummy data for the PTI example. value = value for a hypothetical variable. agg. = spatial aggregation (arithemtic mean); trunc. truncation; stat. = spatial statistics. The transformation step is left out for simplicity. Pathways refer to figure 3."

@@

We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
It should simply work, without the need for elaborate explanations.
Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
This can, however, happen, and has happened, when nested ECA use different aggregation pathways for the same indicator, for example when nested assessments use a different number of aggregation levels.

Consider a case ECA was to be made for a country.
For one of the variables, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
Then in a later regional ECA, the same indicator was estimated for separate regions, as well as for the whole country.
When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation, causing potentially the two nested assessment to reach different conclusions.
A third account, focusing on municipalities (using municipalities as ecosystem assets, rather than the larger regions), will create a third answer to the same question.
We believe it a very desirable property of ECA to be compliant and harmonised assessment scales.
In this specific case, one could in the first national account have performed a step-wise aggregation via municipalities, regions, and finally to the entire EAA, but it would not always be possible to know which ecosystem asset are to be used in future accounts.
We believe one solution to this is to acknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
Fr each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
This would show ECA users if two accounts are using different metrics that cannot be assumed to behave similarly or report the same thing, even if they have the same common name and data source.

@@

```{=html}
<!-- The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
This has to do with the use of area weighting in the spatial aggregation. -->
```

```{=html}
<!-- 
Consider an indicator originating from point-based field surveys.
The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation.
However, if in a later ECA, ecosystem assets are defined differently, e.g. as subsets of the initial assets (a finer scale account), 
then the same indicator, when aggregated to the EAA level, will have a different value. -->
```

@@

<!-- *Pathway*: 1 -->

In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale (similar to @fig-agg1 pane B, but without truncation, as that is not needed in the alien species example).
When performing an ECA, one has the option to either aggregate variable values for each ecosystem asset, and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons (Basal spatial unit (BSU) scale), and then aggregate the indicator values to the regional level (pathway 1).
Arguably the variable and the reference levels are at commensurable scales in both cases (cf. Example 1), and other arguments for choosing one over the other must be found.\

>>>>

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
  - aggregation error
  - aggregation displacement
  - upscaling
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
callout-appearance: simple
callout-icon: false
---

@@

\clearpage

@@

\clearpage

::: callout-note
**Box 1 -- Related concepts**\

@@

It is common in the ECA community of practice to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e. horizontal) aggregation.\

@@

The modifiable area unit problem (MAUP) describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
Administrative units are often used as the ecosystem assets in SEEA EAs, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
The original spatial units that data are collected at may be a pragmatic choice in much the same way.
Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.\

@@

@@

```{r read}
#| include: false
#| cache: false
dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
  mutate(X = X |> str_replace_all("%", "\\\\%"))
cn <- colnames(dat) |>
    str_replace_all("2", "") |>
    str_replace_all("X", "") |>
    str_replace_all("RL", "X\\\\textsubscript{100}")
```

```{r tbl-pti}
#| echo: false
#| tbl-cap: "Dummy data for the PTI example. value = value for a hypothetical variable. agg. = spatial aggregation (arithemtic mean); trunc. truncation; stat. = spatial statistics. The transformation step is left out for simplicity. Pathways refer to figure 3."
#| cache: false

@@

We believe it is a very desirable property of ECAs to be comparable and harmonised across assessment scales, and that accounts from sub-areas should be able to recreate the account of the entire area.
Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
This can happen when nested ECA use different aggregation pathways for the same indicator, for example when nested assessments use a different number of aggregation levels.

Consider a case where an ECA was to be made for a country.
For one of the variables, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
Then in a later regional ECA, the same indicator was estimated for separate regions, as well as for the whole country.
When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation, causing potentially the two nested assessment to reach different conclusions.
A third account, focusing on municipalities (using municipalities as ecosystem assets, rather than the larger regions), will create a third answer to the same question.
In this specific case, one could in the first national account have performed a step-wise aggregation via municipalities, regions, and finally to the entire EAA, but it would not always be possible to know which ecosystem asset are to be used in future accounts.

We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
It should simply work, without the need for elaborate explanations.

We believe one solution to this is to acknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
For each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
This would show ECA users if two accounts are using different metrics that cannot be assumed to behave similarly or report the same thing, even if they have the same common name and data source.

@@

@@

In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale (similar to @fig-agg1 pane B, but without truncation, as that is not needed in this example).
When performing an ECA, one has the option to either aggregate variable values for each ecosystem asset, and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons (Basal spatial unit (BSU) scale), and then aggregate the indicator values to the regional level (pathway 1).
Arguably the variable and the reference levels are at commensurable scales in both cases (cf. Example 1), and other arguments for choosing one over the other must be found.\
